Comparison of Fast, Hybrid Imaging Architectures for Multi-scale, Hierarchical Aperture Arrays

### 2024-10-19 response_to_review_1.txt

--> I thank the reviewer for a thorough review of the manuscript. I have addressed the concerns as detailed below. My responses are preceded with "-->". Changes in the manuscript have been highlighted with bold font. I hope this adequately addresses all the concerns and the manuscript is considered for publication.

Comments to the Author
This paper addresses the important question of how to choose a computational architecture for a given aperture-array radio telescope depending on how its collecting area is deployed, and specifically in the context of some current/upcoming arrays that use hierarchical distribution of antennas. The analysis is new and useful to the community. I found much of the paper to be quite dense in its presentation which unfortunately detracts from its readability. I address one aspect of this in the major comments below.

Major comments
=============
- The paper gets dense (in terms of readability) at the start of Section 5 and I would suggest more care to ensure that the key equations are well described. For example, equation (1) is presented with essentially no context, and the notation is not well explained. Many of the indices are undefined (what is p? what is alpha? what is the asterisk?). This really holds throughout all of Sections 5 and 6. Even though it will lengthen the paper somewhat, I think it’s crucial that you review these approaches and their associated equations carefully, ensuring that they are framed usefully for the reader and that the notation is clarified.

--> Taking this issue of readability very seriously, I have made significant additions to virtually every section and subsection from section 5 onwards. All equations are now described with clearer notation along with the meaning and interpretation between the terms in the equations and the equations themselves. It has increased the manuscript length slightly, but I believe it has significantly enhanced the readability of the manuscript.

### 2024-10-19 ms.tex
Update figure 1 caption, response to referee

The parameters for the chosen examples are summarised in Table~\ref{tab:array_params}. Nominal wavelengths, $\lambda$, at which the arrays are planned to operate are listed. The computational costs are typically independent of the wavelength and only depend on the spatial frequencies as shown in Table~\ref{tab:array_params}. However, the aperture efficiencies can, in general, vary with wavelength. For a fair comparison, the apertures are all assumed to be 100\% efficient at the nominal wavelengths chosen, and that the arrays are used for real-time imaging. \textbf{Note that traditionally cosmological applications such as probing the epoch of reionisation using redshifted 21~cm have relied on saving the visibilities and processing them offline. In such cases, although imaging costs like gridding, Discrete Fourier Transform (DFT), or FFT will not apply, the correlations must still be obtained in real-time. With large arrays, the traditional correlator approach may be too expensive. For example, the Packed Ultra-wideband Mapping Array \citep[PUMA;][]{PUMA+2019}, a planned radio telescope for cosmology and transients with $\sim 5000-32000$ elements, is considering an FFT correlator architecture, a specific adaptation of the EPIC architecture for redundant spacings. This study, with a breakdown of costs for different operations, is also relevant for such cosmological arrays.} Although aperture arrays consisting of only three hierarchical scales are considered here, the analysis can be extended without loss of generality to arrays of even larger levels of spatial hierarchy.

@@

Two broad classes of two-stage fast imaging architectures are considered. In both classes, the first stage is intra-station interferometry of the element electric fields using the options of voltage beamforming (BF), E-field Parallel Imaging Correlator (EPIC), beamforming of correlations (XBF), or correlation followed by FFT (XFFT). The second stage \textbf{involves} inter-station combination of the data products from the first stage. The main difference between the two classes is in whether the second stage uses the first stage products coherently or incoherently. A schematic of these hybrid architectures and their computational cost components are illustrated in Figure~\ref{fig:hybrid-architectures-cost-schematic}.

@@

\begin{figure}
\includegraphics[width=\linewidth]
  % {figures/flops_calc_abridged.pdf}
{fig_01.pdf}
\caption{A hybrid, two-stage architecture for hierarchical aperture arrays. Black arrows indicate pathways where the signal consists of electric fields with phase coherence. Gray arrows indicate pathways of a ``squared'' or a correlated signal with no absolute but only relative phase information. At the top are intra-station architectures, namely, BF, correlator, and EPIC. The top box denotes intra-station imaging options. The bottom box denotes inter-station imaging architectures that process coherent outputs from multiple stations deploying intra-station BF and/or EPIC architectures. 
\label{fig:hybrid-architectures-cost-schematic}}
\end{figure}

>>>>

The parameters for the chosen examples are summarised in Table~\ref{tab:array_params}. Nominal wavelengths, $\lambda$, at which the arrays are planned to operate are listed. The computational costs are typically independent of the wavelength and only depend on the spatial frequencies as shown in Table~\ref{tab:array_params}. However, the aperture efficiencies can, in general, vary with wavelength. For a fair comparison, the apertures are all assumed to be 100\% efficient at the nominal wavelengths chosen, and that the arrays are used for real-time imaging. \textbf{Note that traditionally cosmological applications such as probing the epoch of reionisation using redshifted 21~cm have relied on saving the visibilities and processing them offline. In such cases, although imaging costs like gridding, Discrete Fourier Transform (DFT), or FFT will not apply, the correlations must still be obtained in real-time. With large arrays, the traditional correlator approach may be too expensive. For example, the Packed Ultra-wideband Mapping Array \citep[PUMA;][]{PUMA+2019}, a planned radio telescope for cosmology and transients with $\sim 5000-32000$ elements, is considering an FFT correlator architecture, a specific adaptation of the EPIC architecture for redundant spacings. This study, with a breakdown of costs for different operations, is also relevant for such cosmological arrays.}

@@

Two broad classes of two-stage fast imaging architectures are considered. In both classes, the first stage is intra-station interferometry of the element electric fields using the options of voltage beamforming (BF), E-field Parallel Imaging Correlator (EPIC), beamforming of correlations (XBF), or correlation followed by FFT (XFFT). The second stage \textbf{involves} inter-station combination of the data products from the first stage. The main difference between the two classes is in whether the second stage uses the first stage products coherently or incoherently. \textbf{Although aperture arrays consisting of only three hierarchical scales are considered here, the analysis can be extended without loss of generality to arrays of even larger levels of spatial hierarchy.} A schematic of these hybrid architectures, their computational cost components, \textbf{and their extension to multiple stages} are illustrated in Figure~\ref{fig:hybrid-architectures-cost-schematic}.

@@

\begin{figure}
\includegraphics[width=0.95\linewidth]
  % {figures/flops_calc_abridged_v2.pdf}
{fig_01.pdf}
\caption{A hybrid, two-stage architecture for hierarchical aperture arrays \textbf{showing the methodology for extension to multiple stages.} Black arrows indicate pathways where the signal consists of electric fields with phase coherence. Gray arrows indicate pathways of a ``squared'' or a correlated signal with no absolute but only relative phase information. At the top are intra-station architectures, namely, BF, correlator, and EPIC. The top box denotes intra-station imaging options. The bottom box denotes inter-station imaging architectures that process coherent outputs from multiple stations deploying intra-station BF and/or EPIC architectures. \textbf{The electric fields phase coherently synthesised by BF or EPIC at any stage can form the input for the next stage, thereby allowing the extension of the hierarchy.}
\label{fig:hybrid-architectures-cost-schematic}}
\end{figure}
